\pkg{mlr3spatiotempcv}: Spatiotemporal Resampling Methods for \pkg{mlr3}

### 2021-04-02 paper/paper.Rmd
- use pdf_book reference mechanism - use cache - adjust figure placements - restructure chapters - write CSTF methods

  The goal of [mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com) is to gather the available spatiotemporal resampling methods in R and make them available to users through a simplified and generic interface.
  The latter is made possible by integrating the package directly into the [mlr3](https://mlr3.mlr-org.com) machine learning framework.
  This simplifies the step of integrating package specific syntax into an overarching machine learning pipeline.
  The hope is that the availability of such a package encourages researchers to more often apply spatiotemporal resampling methods in scientific case studies.
keywords:
  # at least one keyword must be supplied
  formatted: [resampling, machine learning, autocorrelation, spatial, temporal, "\\proglang{R}"]
  plain:     [resampling, machine learning, autocorrelation, spatial, temporal, R]
preamble: >
  \usepackage{amsmath}
  \usepackage{booktabs}
  \usepackage{longtable}
tables: true
bibliography: paper.bib
output: rticles::jss_article
---

@@

To introduce the effect of spatial autocorrelation in modeling, a spatial and non-spatial CV using \pkg{mlr3spatiotempcv} will be shown in the following.
For simplicity, one of the most used methods (spatial k-means clustering after @sperrorest) will be used.
Individual methods are discussed in FIXME.

Last, the random forest learner (`lrn("classif.ranger")`) is initialized with default hyperparameters and the prediction type set to `"probability"` due to the binary response variable.
A set of commonly used learners is available in \pkg{mlr3learners} [@mlr3learners], including the random forest implementation after @ranger.

```{r setup, echo=FALSE}
# important for correct figure placement
# see https://stackoverflow.com/a/49395389/4185785
knitr::knit_hooks$set(plot = function(x, options) {
  knitr::hook_plot_tex(x, options)
})
knitr::opts_chunk$set(cache = FALSE)
```

```{r prepare, cache=FALSE}
library("mlr3")
library("mlr3spatiotempcv")
library("mlr3learners")
set.seed(42)

@@

```{r rr-nspcv, cache=FALSE}
resampling_nsp <- rsmp("repeated_cv", folds = 4, repeats = 2)$instantiate(task)
rr_nsp <- resample(
  task = task, learner = learner,
  resampling = resampling_nsp
)

```{r rr-spcv, cache=FALSE}
resampling_sp <- rsmp("repeated_spcv_coords", folds = 4, repeats = 2)$instantiate(task)
rr_sp <- resample(
  task = task, learner = learner,
  resampling = resampling_sp
)

Here, the estimated AUC using Non-Spatial Cross-Validation (NSpCV) is around 0.15 better compared to the Spatial Cross-Validation (SpCV).
The better performance score results from over-optimism in the estimation with NSpCV.
The actual predictive performance on new data is much more realistically reflected by the performance estimated via SpCV.
The magnitude of this difference is variable as it depends on the dataset, the magnitude of STAC and the learner itself.
Algorithms with a higher tendency of overfitting to the training set will have a large spread in such scenarios.

@@

At the time of writing, the following spatiotemporal resampling methods (FIXME ref table) were available across different R packages.
This list might increase over time and it is recommended to consult the homepage of \pkg{mlr3spatiotempcv} for an up-to-date list.

@@

Due to its similarity to a LOO CV, this method will result in as many train/test evaluations as observations are present.
In the example below, a buffer zone of 1000 m is used and the first fold shown is plotted.

```{r}
resampling_buffer <- rsmp("spcv_buffer", theRange = 1000)$instantiate(task)

autoplot(resampling_buffer, size = 0.7, task = task, fold_id = 1, crs = 4326) *
  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
```

@@

`selection = "random"`

```{r}
resampling_block_random <- rsmp("spcv_block", range = 1000, folds = 5)

@@

`selection = "systematic"`: Here, aggregation follows the linear count from $1:n_{folds} from top to bottom.

```{r}
resampling_block_systematic <- rsmp("spcv_block",
  range = 1000, folds = 5,
  selection = "systematic")

@@

`selection = "checkerboard"`: This option ignores paramter `range` and always creates two partitions following a "1,2,1,2" pattern.

```{r}
resampling_block_checkerboard <- rsmp("spcv_block",
  range = 1000, folds = 5,
  selection = "checkerboard")

@@

```{r}
resampling_env <- rsmp("spcv_env",
  features = "distdeforest", folds = 5)

@@

```{r}
resampling_env_multi <- rsmp("spcv_env",
  features = c("distdeforest", "slope"), folds = 5)

@@

```{r}
resampling_coords <- rsmp("spcv_coords", folds = 5)$instantiate(task)

@@

## Leave-Location-and-Time-Out (LLOT)

Leave-Location-and-Time-Out from packge \pkg{CAST} is a spatiotemporal resampling method.
It requires both a spatial and a temporal variable which are used to to separate observations in a multi-dimensional space.

@@

The visualization is done in a dynamic way via \pkg{plotly} to visualize both temporal and spatial dimensions.
Because a dynamic image cannot be included in this manuscript, a static version generated by setting `static_image = TRUE` was used.

```{r}
task_spt = tsk("cookfarm")
resampling_cstf <- rsmp("sptcv_cstf",
  folds = 5,
  time_var = "Date", space_var = "SOURCEID")

autoplot(resampling_cstf,
  fold_id = 1, task = task_spt, crs = 4326, static_image = TRUE,
  file = "plot.pdf") *
  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
```

```{r, echo=FALSE}
knitr::include_graphics("plot.pdf", dpi = 100)
```

@@

##

Every partitioning method in [mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com) comes with a generic `plot()` method to visualize the created groups.
In a 2D space this happens via [ggplot2](https://ggplot2.tidyverse.org) [@ggplot2] while for spatiotemporal methods 3D visualizations via [plotly](https://github.com/ropensci/plotly) [@plotly] are created.

```{r vis-spcv, fig.cap="Visualization of spatial partitioning", fig.pos="!b", fig.height=6, fig.width=8}
autoplot(resampling_sp, task, fold_id = c(1:4)) *
  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
```
\

This example used an already created task via the convenience function `tsk()`.
In practice however, one needs to create a spatiotemporal task via `TaskClassifST()`/`TaskRegrST()` and set the `crs` argument.

The spatial grouping of the k-means based approach above (Figure 1) contrasts visually very well compared to the NSpCV (random) partitioning (Figure 2):

```{r vis-nspcv, fig.cap="Visualization of random partitioning", fig.pos="!b", fig.height=6, fig.width=8}
autoplot(resampling_nsp, task, fold_id = c(1:4), crs = 4326) *
  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
```

>>>>

  The goal of [mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com) is to gather the available spatiotemporal resampling methods in R and make them available to users through a simplified and generic interface.
  The latter is made possible by integrating the package directly into the [mlr3](https://mlr3.mlr-org.com) machine learning framework.
  This simplifies the step of integrating package specific syntax into an overarching machine learning pipeline.
  The hope is that the availability of such a package encourages researchers to more often apply spatiotemporal resampling methods in scientific case studies.
keywords:
  # at least one keyword must be supplied
  formatted: [resampling, machine learning, autocorrelation, spatial, temporal, "\\proglang{R}"]
  plain:     [resampling, machine learning, autocorrelation, spatial, temporal, R]
preamble: >
  \usepackage{amsmath}
  \usepackage{booktabs}
  \usepackage{longtable}
tables: true
bibliography: paper.bib
# output: rticles::jss_article
output:
  bookdown::pdf_book:
    base_format: rticles::jss_article
---

```{r setup, echo=FALSE}
# important for correct figure placement
# see https://stackoverflow.com/a/49395389/4185785
knitr::knit_hooks$set(plot = function(x, options) {
  knitr::hook_plot_tex(x, options)
})
# knitr::opts_chunk$set(cache = FALSE)
```

@@

Every partitioning method in [mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com) comes with a generic `plot()` method to visualize the created groups.
In a 2D space this happens via [ggplot2](https://ggplot2.tidyverse.org) [@ggplot2] while for spatiotemporal methods 3D visualizations are created via [plotly](https://github.com/ropensci/plotly) [@plotly].

Note: The following example uses a built-in task via function `tsk()`.
In practice however, one needs to create a spatiotemporal task first via `TaskClassifST()`/`TaskRegrST()`..

The spatial grouping of the k-means based approach (`"spcv_coords"`) (Fig.\@ref(fig:vis-spcv)) contrasts visually very well to the NSpCV (random) partitioning (Fig.\@ref(fig:vis-nspcv)).

```{r vis-spcv, fig.cap="Visualization of spatial partitioning", fig.pos="t!", fig.height=6, fig.width=8, cache=TRUE}
library(mlr3)
library(mlr3spatiotempcv)

@@

set.seed(42)

task = tsk("ecuador")

resampling_sp <- rsmp("repeated_spcv_coords", folds = 4, repeats = 2)
resampling_sp$instantiate(task)

autoplot(resampling_sp, task, fold_id = c(1:2), crs = 4326, size = 0.8) *
  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
```

```{r vis-nspcv, fig.cap="Visualization of random partitioning", fig.pos="t!", fig.height=6, fig.width=8, cache=T}
resampling_nsp <- rsmp("repeated_cv", folds = 4, repeats = 2)
resampling_nsp$instantiate(task)

autoplot(resampling_nsp, task, fold_id = c(1:2), crs = 4326, size = 0.8) *
  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
```

@@

At the time of writing, the following spatiotemporal resampling methods (Table \@ref(tab:sptcv-methods)) were available across different R packages.
This list might increase over time and it is recommended to consult the homepage of \pkg{mlr3spatiotempcv} for an up-to-date list.

@@

Due to its similarity to a LOO CV, this method will result in as many train/test evaluations as observations are present.
In the example below, a buffer zone of 1000 m is used and the first fold shown is plotted (Fig.\@ref(fig:buffer))).

```{r buffer, fig.cap="Visualization of 'Spatial buffering' method.", out.width="50%", cache=T}
resampling_buffer <- rsmp("spcv_buffer", theRange = 1000)
resampling_buffer$instantiate(task)

autoplot(resampling_buffer,
  size = 0.8, task = task, fold_id = 1, crs = 4326) *
  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
```

@@

### Option: selection = "random"

```{r block-random, cache = TRUE, fig.cap="Spatial blocking with option `selection = 'random'`.", out.width="50%", fig.pos="h"}
resampling_block_random <- rsmp("spcv_block", range = 1000, folds = 5)

@@

### Option: selection = "systematic"

Here, aggregation follows the linear count from $1:n_{folds}$ from top to bottom.

```{r block-systematic, cache=TRUE, fig.cap="Spatial blocking with option `selection = 'systematic'`.", out.width="50%", fig.pos="h"}
resampling_block_systematic <- rsmp("spcv_block",
  range = 1000, folds = 5,
  selection = "systematic"
)

@@

### Option: selection = "checkerboard"

This option ignores parameter `range` and always creates two partitions following a "1,2,1,2" pattern.

```{r block-checkerboard, cache=TRUE, fig.cap="Spatial blocking with option `selection = 'checkerboard'`.", out.width="50%", fig.pos="h"}
resampling_block_checkerboard <- rsmp("spcv_block",
  range = 1000, folds = 5,
  selection = "checkerboard")

@@

```{r env-1, fig.cap="Environmental blocking using a single predictor.", cache=TRUE, out.width="50%", fig.pos="h"}
resampling_env <- rsmp("spcv_env",
  features = "distdeforest", folds = 5)

@@

```{r env-2, fig.cap="Environmental blocking using multiple predictors.", cache=TRUE, out.width="50%", fig.pos="h"}
resampling_env_multi <- rsmp("spcv_env",
  features = c("distdeforest", "slope"), folds = 5)

@@

```{r coords, fig.cap="Method 'Spatial CV' showing a k-means clustering approach based on coordinates.", cache=TRUE, out.width="50%", fig.pos="h"}
resampling_coords <- rsmp("spcv_coords", folds = 5)$instantiate(task)

@@

## CreateSpacetimeFolds (Cstf)

Leave-Location-and-Time-Out from packge \pkg{CAST} is a spatiotemporal resampling method [@meyer2018a].
It requires both a spatial and a temporal variable which are used to to separate observations in a multi-dimensional space.
It comes with satellite methods named "Leave-time-out" (LTO) and "Leave-location-out" (LLO) which only operatre in a two-dimensional space.

@@

Two visualization types are supported: a two-dimensional and a three-dimensional method which can be selected via the logical argument `plot3d`.
The 3D visualization has been implemented via \pkg{plotly}.
Because a dynamic image cannot be included in this manuscript, a static version, which can be generated by setting `static_image = TRUE`, is shown.

In \pkg{mlr3spatiotempcv} the method is named `"sptcv_cstf"` because the underlying upstream implementation in package \pkg{CAST} is named `CreateSpacetimeFolds()`.

### Leave-Time-Out (LTO)

To better showcase the partitioning of the "Leave-Time-Out" (LTO) method, the temporal variable `Date` was adjusted in this example.

The following code uses the `cookfarm` dataset as the base, modifies to the `Date` variable to contain five unique levels and then creates a spatiotemporal regression task in \pkg{mlr3spatiotempcv} (Fig.\@ref(fig:lto)).

```{r, results='hide', cache=TRUE}
data = cookfarm_sample
data$Date = rep(c(
  "2020-01-01", "2020-02-01", "2020-03-01", "2020-04-01",
  "2020-05-01"), times = 1, each = 100)
b = mlr3::as_data_backend(data)
b$hash = "_mlr3_tasks_cookfarm_"
task_spt = TaskRegrST$new(
  id = "cookfarm", b, target = "PHIHOX",
  extra_args = list(
    coordinate_names = c("x", "y"), coords_as_features = FALSE,
    crs = 26911)
)

resampling_cstf_time = rsmp("sptcv_cstf", folds = 5, time_var = "Date")
resampling_cstf_time$instantiate(task_spt)

autoplot(resampling_cstf_time,
  fold_id = 5, task = task_spt, crs = 4326, plot3D = TRUE
)
```

```{r lto, echo=FALSE, out.width="90%", fig.cap="Three-dimensional visualization of method `sptcv\\_cstf` and sub-method Leave-Time-Out (LTO). To support showcasing the temporal grouping, five folds and five temporal levels in variable `Date` were used in this example.", fig.pos="h"}
knitr::include_graphics("lto.png")
```

### Leave-Location-Out (LLO)

Method "Leave-Location-Out" (LLO) works similar to "Leave-Time-Out" (LTO) in that it operates in a two-dimensional space.
Here, specific locations are left out for forming the partitions while the temporal information is ignored (Fig.\@ref(fig:llo)).

```{r cache=TRUE, results='hide'}
resampling_cstf_loc = rsmp("sptcv_cstf", folds = 5, space_var = "SOURCEID")
resampling_cstf_loc$instantiate(task_spt)

autoplot(resampling_cstf_loc,
  fold_id = 5, task = task_spt, crs = 4326, plot3D = TRUE)
```

```{r llo, echo=FALSE, out.width="90%", fig.cap="Three-dimensional visualization of method 'sptcv\\_cstf' and sub-method 'Leave-Location-Out' (LLO). To support showcasing the spatial grouping, the viewing angle has been adjusted to a view from top. Points are stacked by temporal levels and only points from the same location are chosen to form a partition.", fig.pos="h"}
knitr::include_graphics("llo.png")
```

### Leave-Location-and-Time-Out (LLTO)

The "Leave-Location-and-Time-Out" (LLTO) method combines both "Leave-Time-Out" (LTO) and "Leave-Location-Out" (LLO).
It creates partitions that operate on spatiotemporal clusters to reduce both spatial and temporal autocorrelations.
To achieve this, observations are omitted and only a subset is used.
In the special scenario with five temporal levels and five folds, the train/test split happens only at one temporal level (Fig.\@ref(fig:llto)).

```{r cache=TRUE, results='hide'}
resampling_cstf_time_loc = rsmp("sptcv_cstf",
  folds = 5,
  space_var = "SOURCEID", time_var = "Date")
resampling_cstf_time_loc$instantiate(task_spt)

autoplot(resampling_cstf_time_loc,
  fold_id = 4, task = task_spt, crs = 4326, plot3D = TRUE,
  show_omitted = TRUE)
```

```{r llto, echo=FALSE, out.width="90%", fig.cap="Three-dimensional visualization of method 'sptcv\\_cstf' and sub-method 'Leave-Location-And-Time-Out' (LLTO). The grey points show the points omitted by this method.", cache=TRUE, fig.pos="h"}
knitr::include_graphics("llto.png")
```

@@

To show the effect of spatial autocorrelation in modeling, a spatial and non-spatial CV using \pkg{mlr3spatiotempcv} will be shown in the following.
For simplicity, one of the most used methods (spatial k-means clustering after @sperrorest) will be used.
Individual methods are discussed in FIXME.

Next, the random forest learner (`lrn("classif.ranger")`) is initialized with default hyperparameters and the prediction type set to `"probability"` due to the binary response variable.
A set of commonly used learners is available in \pkg{mlr3learners} [@mlr3learners], including the random forest implementation after @ranger.

```{r prepare, cache=TRUE}
library("mlr3")
library("mlr3spatiotempcv")
library("mlr3learners")

Here, a four-fold, two times repeated non-spatial resampling structure is constructed and instantianeted via method `$instantiate()`.
Next, the created resampling object `resampling_nsp` is passed to function `resample()` together with the Task and Learner created earlier to execute the cross-validation.

Last, the results from the CV are aggregated using the measure `"classif.auc"`, which is the technical notation for the Area Under the Receivers Operating Curve (AUROC).

```{r rr-nspcv, cache=TRUE, cache.lazy = FALSE}
resampling_nsp <- rsmp("repeated_cv", folds = 4, repeats = 2)
resampling_nsp$instantiate(task)
rr_nsp <- resample(
  task = task, learner = learner,
  resampling = resampling_nsp
)

The same approach as for non-spatial resampling is used for spatial resampling, expect the method being `"repeated_spcv_coords"` instead of `"repeated_cv"`.

```{r rr-spcv, cache=TRUE, cache.lazy = FALSE}
resampling_sp <- rsmp("repeated_spcv_coords", folds = 4, repeats = 2)
resampling_sp$instantiate(task)
rr_sp <- resample(
  task = task, learner = learner,
  resampling = resampling_sp
)

The resulting AUC using Non-Spatial Cross-Validation (NSpCV) is around 0.15 better compared to the Spatial Cross-Validation (SpCV).
The better performance score results from over-optimism in the estimation with NSpCV.
The actual predictive performance on new data is much more realistically reflected by the performance estimated via SpCV.
The magnitude of this difference is variable as it depends on the dataset, the magnitude of STAC and the learner itself.
Algorithms with a higher tendency of overfitting to the training set will have a large spread in such scenarios.
